\chapter{}
\begin{enumerate}
\item The great sages approached Manu, who was seated with a collected mind, and, having duly worshipped him, spoke as follows:
\item `Deign, divine one, to declare to us precisely and in due order the sacred laws of each of the (four chief) castes (varna) and of the intermediate ones.
\item `For thou, O Lord, alone knowest the purport, (i.e.) the rites, and the knowledge of the soul, (taught) in this whole ordinance of the Self-existent (Svayambhu), which is unknowable and unfathomable.'
\item He, whose power is measureless, being thus asked by the high-minded great sages, duly honoured them, and answered, `Listen!'
\item This (universe) existed in the shape of Darkness, unperceived, destitute of distinctive marks, unattainable by reasoning, unknowable, wholly immersed, as it were, in deep sleep.
\item Then the divine Self-existent (Svayambhu, himself) indiscernible, (but) making (all) this, the great elements and the rest, discernible, appeared with irresistible (creative) power, dispelling the darkness.
\item He who can be perceived by the internal organ (alone), who is subtile, indiscernible, and eternal, who contains all created beings and is inconceivable, shone forth of his own (will).
\item He, desiring to produce beings of many kinds from his own body, first with a thought created the waters, and placed his seed in them.
\item That (seed) became a golden egg, in brilliancy equal to the sun; in that (egg) he himself was born as Brahman, the progenitor of the whole world.
\item The waters are called narah, (for) the waters are, indeed, the offspring of Nara; as they were his first residence (ayana), he thence is named Narayana.
\item From that (first) cause, which is indiscernible, eternal, and both real and unreal, was produced that male (Purusha), who is famed in this world (under the appellation of) Brahman.
\item The divine one resided in that egg during a whole year, then he himself by his thought (alone) divided it into two halves;
\item And out of those two halves he formed heaven and earth, between them the middle sphere, the eight points of the horizon, and the eternal abode of the waters.
\item From himself (atmanah) he also drew forth the mind, which is both real and unreal, likewise from the mind egoism, which possesses the function of self-consciousness (and is) lordly;
\item Moreover, the great one, the soul, and all (products) affected by the three qualities, and, in their order, the five organs which perceive the objects of sensation.
\item But, joining minute particles even of those six, which possess measureless power, with particles of himself, he created all beings.
\item Because those six (kinds of) minute particles, which form the (creator's) frame, enter (a-sri) these (creatures), therefore the wise call his frame sarira, (the body.)
\item That the great elements enter, together with their functions and the mind, through its minute parts the framer of all beings, the imperishable one.
\item But from minute body (-framing) particles of these seven very powerful Purushas springs this (world), the perishable from the imperishable.
\item Among them each succeeding (element) acquires the quality of the preceding one, and whatever place (in the sequence) each of them occupies, even so many qualities it is declared to possess.
\item But in the beginning he assigned their several names, actions, and conditions to all (created beings), even according to the words of the Veda.
\item He, the Lord, also created the class of the gods, who are endowed with life, and whose nature is action; and the subtile class of the Sadhyas, and the eternal sacrifice.
\item But from fire, wind, and the sun he drew forth the threefold eternal Veda, called Rik, Yagus, and Saman, for the due performance of the sacrifice.
\item Time and the divisions of time, the lunar mansions and the planets, the rivers, the oceans, the mountains, plains, and uneven ground.
\item Austerity, speech, pleasure, desire, and anger, this whole creation he likewise produced, as he desired to call these beings into existence.
\item Moreover, in order to distinguish actions, he separated merit from demerit, and he caused the creatures to be affected by the pairs (of opposites), such as pain and pleasure.
\item But with the minute perishable particles of the five (elements) which have been mentioned, this whole (world) is framed in due order.
\item But to whatever course of action the Lord at first appointed each (kind of beings), that alone it has spontaneously adopted in each succeeding creation.
\item Whatever he assigned to each at the (first) creation, noxiousness or harmlessness, gentleness or ferocity, virtue or sin, truth or falsehood, that clung (afterwards) spontaneously to it.
\item As at the change of the seasons each season of its own accord assumes its distinctive marks, even so corporeal beings (resume in new births) their (appointed) course of action.
\item But for the sake of the prosperity of the worlds he caused the Brahmana, the Kshatriya, the Vaisya, and the Sudra to proceed from his mouth, his arms, his thighs, and his feet.
\item Dividing his own body, the Lord became half male and half female; with that (female) he produced Virag.
\item But know me, O most holy among the twice-born, to be the creator of this whole (world), whom that male, Virag, himself produced, having performed austerities.
\item Then I, desiring to produce created beings, performed very difficult austerities, and (thereby) called into existence ten great sages, lords of created beings,
\item Mariki, Atri, Angiras, Pulastya, Pulaha, Kratu, Praketas, Vasishtha, Bhrigu, and Narada.
\item They created seven other Manus possessing great brilliancy, gods and classes of gods and great sages of measureless power,
\item Yakshas (the servants of Kubera, the demons called) Rakshasas and Pisakas, Gandharvas (or musicians of the gods), Apsarases (the dancers of the gods), Asuras, (the snake-deities called) Nagas and Sarpas, (the bird-deities called) Suparnas and the several classes of the manes,
\item Lightnings, thunderbolts and clouds, imperfect (rohita) and perfect rainbows, falling meteors, supernatural noises, comets, and heavenly lights of many kinds,
\item (Horse-faced) Kinnaras, monkeys, fishes, birds of many kinds, cattle, deer, men, and carnivorous beasts with two rows of teeth,
\item Small and large worms and beetles, moths, lice, flies, bugs, all stinging and biting insects and the several kinds of immovable things.
\item Thus was this whole (creation), both the immovable and the movable, produced by those high-minded ones by means of austerities and at my command, (each being) according to (the results of) its actions.
\item But whatever act is stated (to belong) to (each of) those creatures here below, that I will truly declare to you, as well as their order in respect to birth.
\item Cattle, deer, carnivorous beasts with two rows of teeth, Rakshasas, Pisakas, and men are born from the womb.
\item From eggs are born birds, snakes, crocodiles, fishes, tortoises, as well as similar terrestrial and aquatic (animals).
\item From hot moisture spring stinging and biting insects, lice, flies, bugs, and all other (creatures) of that kind which are produced by heat.
\item All plants, propagated by seed or by slips, grow from shoots; annual plants (are those) which, bearing many flowers and fruits, perish after the ripening of their fruit;
\item (Those trees) which bear fruit without flowers are called vanaspati (lords of the forest); but those which bear both flowers and fruit are called vriksha.
\item But the various plants with many stalks, growing from one or several roots, the different kinds of grasses, the climbing plants and the creepers spring all from seed or from slips.
\item These (plants) which are surrounded by multiform Darkness, the result of their acts (in former existences), possess internal consciousness and experience pleasure and pain.
\item The (various) conditions in this always terrible and constantly changing circle of births and deaths to which created beings are subject, are stated to begin with (that of) Brahman, and to end with (that of) these (just mentioned immovable creatures).
\item When he whose power is incomprehensible, had thus produced the universe and men, he disappeared in himself, repeatedly suppressing one period by means of the other.
\item When that divine one wakes, then this world stirs; when he slumbers tranquilly, then the universe sinks to sleep.
\item But when he reposes in calm sleep, the corporeal beings whose nature is action, desist from their actions and mind becomes inert.
\item When they are absorbed all at once in that great soul, then he who is the soul of all beings sweetly slumbers, free from all care and occupation.
\item When this (soul) has entered darkness, it remains for a long time united with the organs (of sensation), but performs not its functions; it then leaves the corporeal frame.
\item When, being clothed with minute particles (only), it enters into vegetable or animal seed, it then assumes, united (with the fine body), a (new) corporeal frame.
\item Thus he, the imperishable one, by (alternately) waking and slumbering, incessantly revivifies and destroys this whole movable and immovable (creation).
\item But he having composed these Institutes (of the sacred law), himself taught them, according to the rule, to me alone in the beginning; next I (taught them) to Mariki and the other sages.
\item Bhrigu, here, will fully recite to you these Institutes; for that sage learned the whole in its entirety from me.
\item Then that great sage Bhrigu, being thus addressed by Manu, spoke, pleased in his heart, to all the sages, `Listen!'
\item Six other high-minded, very powerful Manus, who belong to the race of this Manu, the descendant of the Self-existent (Svayambhu), and who have severally produced created beings,
\item (Are) Svarokisha, Auttami, Tamasa, Raivata, Kakshusha, possessing great lustre, and the son of Vivasvat.
\item These seven very glorious Manus, the first among whom is Svayambhuva, produced and protected this whole movable and immovable (creation), each during the period (allotted to him).
\item Eighteen nimeshas (twinklings of the eye, are one kashtha), thirty kashthas one kala, thirty kalas one muhurta, and as many (muhurtas) one day and night.
\item The sun divides days and nights, both human and divine, the night (being intended) for the repose of created beings and the day for exertion.
\item A month is a day and a night of the manes, but the division is according to fortnights. The dark (fortnight) is their day for active exertion, the bright (fortnight) their night for sleep.
\item A year is a day and a night of the gods; their division is (as follows): the half year during which the sun progresses to the north will be the day, that during which it goes southwards the night.
\item But hear now the brief (description of) the duration of a night and a day of Brahman and of the several ages (of the world, yuga) according to their order.
\item They declare that the Krita age (consists of) four thousand years (of the gods); the twilight preceding it consists of as many hundreds, and the twilight following it of the same number.
\item In the other three ages with their twilights preceding and following, the thousands and hundreds are diminished by one (in each).
\item These twelve thousand (years) which thus have been just mentioned as the total of four (human) ages, are called one age of the gods.
\item But know that the sum of one thousand ages of the gods (makes) one day of Brahman, and that his night has the same length.
\item Those (only, who) know that the holy day of Brahman, indeed, ends after (the completion of) one thousand ages (of the gods) and that his night lasts as long, (are really) men acquainted with (the length of) days and nights.
\item At the end of that day and night he who was asleep, awakes and, after awaking, creates mind, which is both real and unreal.
\item Mind, impelled by (Brahman's) desire to create, performs the work of creation by modifying itself, thence ether is produced; they declare that sound is the quality of the latter.
\item But from ether, modifying itself, springs the pure, powerful wind, the vehicle of all perfumes; that is held to possess the quality of touch.
\item Next from wind modifying itself, proceeds the brilliant light, which illuminates and dispels darkness; that is declared to possess the quality of colour;
\item And from light, modifying itself, (is produced) water, possessing the quality of taste, from water earth which has the quality of smell; such is the creation in the beginning.
\item The before-mentioned age of the gods, (or) twelve thousand (of their years), being multiplied by seventy-one, (constitutes what) is here named the period of a Manu (Manvantara).
\item The Manvantaras, the creations and destructions (of the world, are) numberless; sporting, as it were, Brahman repeats this again and again.
\item In the Krita age Dharma is four-footed and entire, and (so is) Truth; nor does any gain accrue to men by unrighteousness.
\item In the other (three ages), by reason of (unjust) gains (agama), Dharma is deprived successively of one foot, and through (the prevalence of) theft, falsehood, and fraud the merit (gained by men) is diminished by one fourth (in each).
\item (Men are) free from disease, accomplish all their aims, and live four hundred years in the Krita age, but in the Treta and (in each of) the succeeding (ages) their life is lessened by one quarter.
\item The life of mortals, mentioned in the Veda, the desired results of sacrificial rites and the (supernatural) power of embodied (spirits) are fruits proportioned among men according to (the character of) the age.
\item One set of duties (is prescribed) for men in the Krita age, different ones in the Treta and in the Dvapara, and (again) another (set) in the Kali, in a proportion as (those) ages decrease in length.
\item In the Krita age the chief (virtue) is declared to be (the performance of) austerities, in the Treta (divine) knowledge, in the Dvapara (the performance of) sacrifices, in the Kali liberality alone.
\item But in order to protect this universe He, the most resplendent one, assigned separate (duties and) occupations to those who sprang from his mouth, arms, thighs, and feet.
\item To Brahmanas he assigned teaching and studying (the Veda), sacrificing for their own benefit and for others, giving and accepting (of alms).
\item The Kshatriya he commanded to protect the people, to bestow gifts, to offer sacrifices, to study (the Veda), and to abstain from attaching himself to sensual pleasures;
\item The Vaisya to tend cattle, to bestow gifts, to offer sacrifices, to study (the Veda), to trade, to lend money, and to cultivate land.
\item One occupation only the lord prescribed to the Sudra, to serve meekly even these (other) three castes.
\item Man is stated to be purer above the navel (than below); hence the Self-existent (Svayambhu) has declared the purest (part) of him (to be) his mouth.
\item As the Brahmana sprang from (Brahman's) mouth, as he was the first-born, and as he possesses the Veda, he is by right the lord of this whole creation.
\item For the Self-existent (Svayambhu), having performed austerities, produced him first from his own mouth, in order that the offerings might be conveyed to the gods and manes and that this universe might be preserved.
\item What created being can surpass him, through whose mouth the gods continually consume the sacrificial viands and the manes the offerings to the dead?
\item Of created beings the most excellent are said to be those which are animated; of the animated, those which subsist by intelligence; of the intelligent, mankind; and of men, the Brahmanas;
\item Of Brahmanas, those learned (in the Veda); of the learned, those who recognise (the necessity and the manner of performing the prescribed duties); of those who possess this knowledge, those who perform them; of the performers, those who know the Brahman.
\item The very birth of a Brahmana is an eternal incarnation of the sacred law; for he is born to (fulfil) the sacred law, and becomes one with Brahman.
\item A Brahmana, coming into existence, is born as the highest on earth, the lord of all created beings, for the protection of the treasury of the law.
\item Whatever exists in the world is, the property of the Brahmana; on account of the excellence of his origin The Brahmana is, indeed, entitled to all.
\item The Brahmana eats but his own food, wears but his own apparel, bestows but his own in alms; other mortals subsist through the benevolence of the Brahmana.
\item In order to clearly settle his duties those of the other (castes) according to their order, wise Manu sprung from the Self-existent, composed these Institutes (of the sacred Law).
\item A learned Brahmana must carefully study them, and he must duly instruct his pupils in them, but nobody else (shall do it).
\item A Brahmana who studies these Institutes (and) faithfully fulfils the duties (prescribed therein), is never tainted by sins, arising from thoughts, words, or deeds.
\item He sanctifies any company (which he may enter), seven ancestors and seven descendants, and he alone deserves (to possess) this whole earth.
\item (To study) this (work) is the best means of securing welfare, it increases understanding, it procures fame and long life, it (leads to) supreme bliss.
\item In this (work) the sacred law has been fully stated as well as the good and bad qualities of (human) actions and the immemorial rule of conduct, (to be followed) by all the four castes (varna).
\item The rule of conduct is transcendent law, whether it be taught in the revealed texts or in the sacred tradition; hence a twice-born man who possesses regard for himself, should be always careful to (follow) it.
\item A Brahmana who departs from the rule of conduct, does not reap the fruit of the Veda, but he who duly follows it, will obtain the full reward.
\item The sages who saw that the sacred law is thus grounded on the rule of conduct, have taken good conduct to be the most excellent root of all austerity.
\item The creation of the universe, the rule of the sacraments, the ordinances of studentship, and the respectful behaviour (towards Gurus), the most excellent rule of bathing (on return from the teacher's house),
\item (The law of) marriage and the description of the (various) marriage-rites, the regulations for the great sacrifices and the eternal rule of the funeral sacrifices,
\item The description of the modes of (gaining) subsistence and the duties of a Snataka, (the rules regarding) lawful and forbidden food, the purification of men and of things,
\item The laws concerning women, (the law) of hermits, (the manner of gaining) final emancipation and (of) renouncing the world, the whole duty of a king and the manner of deciding lawsuits,
\item The rules for the examination of witnesses, the laws concerning husband and wife, the law of (inheritance and) division, (the law concerning) gambling and the removal of (men nocuous like) thorns,
\item (The law concerning) the behaviour of Vaisyas and Sudras, the origin of the mixed castes, the law for all castes in times of distress and the law of penances,
\item The threefold course of transmigrations, the result of (good or bad) actions, (the manner of attaining) supreme bliss and the examination of the good and bad qualities of actions,
\item The primeval laws of countries, of castes (gati), of families, and the rules concerning heretics and companies (of traders and the like)- (all that) Manu has declared in these Institutes.
\item As Manu, in reply to my questions, formerly promulgated these Institutes, even so learn ye also the (whole work) from me.
\end{enumerate}
